---
title: "Bayes - Diabetes"
output: 
  word_document:
    toc: false
    reference_docx: "output_template.docx"
    df_print: "kable"
date: "Version `r format(Sys.time(), '%Y-%m-%d')`"

---


```{r setup, echo=FALSE, include=FALSE}
library(tidyverse)
library(brms)
library(bayesplot)
options(mc.cores = 4, brms.backend = "cmdstanr")
devtools::load_all()
theme_set(theme_minimal())
knitr::opts_chunk$set(echo = FALSE)

hmc_iters <- 7000
do_ppc_checks <- FALSE
do_loo <- FALSE
```


```{r}
clinical_data <- read_csv(here::here("private_data/clinical_md_for_bayes.csv"),
                          col_types = cols(
  Patient_ID = col_integer(),
  Time = col_character(),
  bmi_man = col_double(),
  hba1c_man = col_double(),
  ph_man = col_double(),
  dq2_8 = col_character(),
  risk_hla = col_double(),
  c_peptide_change = col_double(),
  c_peptide_change_perc = col_double(),
  gad_man = col_double(),
  iaa_man = col_double(),
  ia2_man = col_double(),
  znt8_man = col_double(),
  idaa1c = col_double(),
  wbc = col_double(),
  lymph_abs = col_double(),
  mon_abs = col_double(),
  neutr_abs = col_double(),
  eoz_abs = col_double(),
  baz_abs = col_double(),
  lymph_perc = col_double(),
  mon_perc = col_double(),
  neutr_perc = col_double(),
  eoz_perc = col_double(),
  baz_perc = col_double(),
  tdd_per_kg_1 = col_double()
))
```


```{r}
cd4_all_with_preliminary <-
  read_csv(
    #here::here("private_data/all_levels_counts_cd4.csv"),
    here::here("private_data/all_levels_counts_with_preliminary_cd4.csv"),
    col_types = cols(
      ...1 = col_skip(),
      Sample_ID = col_character(),
      annotations = col_character(),
      n = col_integer(),
      Condition = col_character(),
      Condition2 = col_character(),
      Disease = col_character(),
      Sex = col_character(),
      Age = col_double(),
      Age_group = col_character(),
      Patient_ID = col_integer(),
      Time = col_character(),
      Experiment_ID = col_character(),
      Level = col_character()
    ) 
  ) %>%
  separate(annotations, into = c("L1", "L2", "L3", "Marker_genes"), sep = "---|: ", fill = "right") %>%
  mutate(Name = case_when(!is.na(L3) ~ L3,
                          !is.na(L2) ~ L2,
                          TRUE ~ L1),
         isPreliminary = Experiment_ID %in% c("Exp08", "Exp10", "Exp11"))

cd4_all <- cd4_all_with_preliminary %>% filter(!isPreliminary) %>% select(-isPreliminary)
```


```{r}
cd4_all_prop_with_preliminary <- cd4_all_with_preliminary %>% 
  group_by(Sample_ID, Level) %>%
  mutate(sum_parent = if_else(Level == "L1", sum(n), NA_real_),
         sum_all = sum(n)) %>%
  group_by(Sample_ID, L1, Level) %>%
  mutate(sum_parent = if_else(Level == "L2", sum(n), sum_parent)) %>%
  group_by(Sample_ID, L1, L2, Level) %>%
  mutate(sum_parent = if_else(Level == "L3", sum(n), sum_parent)) %>%
  ungroup() %>%
  mutate(prop_parent = n / sum_parent,
         prop_all = n / sum_all,
         parent_level = case_when(Level == "L1" ~ "All",
                                  Level == "L2" ~ L1,
                                  Level == "L3" ~ L2))

stopifnot(all(!is.na(cd4_all_prop_with_preliminary$sum_parent)))
stopifnot(all(!is.na(cd4_all_prop_with_preliminary$parent_level)))

prop_sums_bad <- cd4_all_prop_with_preliminary %>%
  group_by(Sample_ID, Level, parent_level) %>%
  summarise(prop_sum = sum(prop_parent), .groups = "drop") %>%
  filter(abs(prop_sum - 1) > 1e-5)

stopifnot(nrow(prop_sums_bad) == 0)

sums_all_bad <- cd4_all_prop_with_preliminary %>%
  group_by(Sample_ID, sum_all) %>%
  tally() %>%
  group_by(Sample_ID) %>%
  filter(n() > 1)

stopifnot(nrow(sums_all_bad) == 0)

cd4_all_prop <- cd4_all_prop_with_preliminary %>% filter(!isPreliminary) %>% select(-isPreliminary)

```

## CD4

```{r}
plot_prop <- function(all_prop, prop_col = prop_parent) {
  all_prop %>%
    filter(!is.na({{prop_col}})) %>%
    ggplot(aes(x = Condition, y = {{prop_col}})) +
    geom_point(position = position_jitter(width = 0.3, height = 0), alpha = 0.5, color = "darkgray") +
    stat_summary(fun.data = "mean_se") +
    facet_wrap(~Name, scales = "free_y")
  
}
cd4_all_prop %>% filter(Level == "L1") %>% plot_prop()
cd4_all_prop %>% filter(Level == "L2" & L2 != "Unconventional") %>% plot_prop()
```


```{r, fig.width=8, fig.height=8}
cd4_all_prop %>% filter(Level == "L3") %>% plot_prop()
```

```{r}
cd4_all_prop %>% filter(Level == "L1") %>% plot_prop(prop_all)
cd4_all_prop %>% filter(Level == "L2" & L2 != "Unconventional") %>% plot_prop(prop_all)
```

```{r}
#cd4_all_prop %>% filter(Level == "L2" & L2 != "Unconventional") %>% 
  
```



```{r, fig.width=8, fig.height=8}
cd4_all_prop %>% filter(Level == "L3") %>% plot_prop(prop_all)
```

Samples with extreme proportions of ISAGhi:

```{r}
df_for_reading <- function(df) {
  df %>% select(Patient_ID, Sample_ID, L2, L3, Condition, Experiment_ID, n, prop_parent, prop_all) %>% mutate(across(starts_with("prop"), scales::percent)) %>%
    rename(Pat. = Patient_ID, Samp. = Sample_ID, Cond. = Condition, Exp = Experiment_ID, `% parent` = prop_parent, `% all` = prop_all)
}

cd4_all_prop %>% filter(Name == "ISAGhi" & prop_parent > 0.1) %>% df_for_reading() %>% select(-L3)
```


```{r}
cd4_l3 <- cd4_all_prop %>%
  filter(Level == "L3") %>%
  select(-Level) 
```


```{r cells-per-sample, fig.cap="Distribution of cells per sample in the CD4 data"}
cd4_l3 %>% group_by(Sample_ID) %>%
  summarise(n_cells = sum(n)) %>%
  ggplot(aes(n_cells)) + geom_histogram(binwidth = 100)
```

### Recalibrating the naive population

```{r}
cd4_preliminary_comparison <- cd4_all_prop_with_preliminary %>% 
  mutate(isPreliminary = if_else(isPreliminary, "preliminary", "final")) %>%
  pivot_wider(names_from = isPreliminary, values_from = c("Sample_ID", "Experiment_ID", "n", "sum_parent", "sum_all", "prop_parent", "prop_all")) %>%
  filter(!is.na(Sample_ID_preliminary))

stopifnot(!any(is.na(cd4_preliminary_comparison$Sample_ID_final)))
#stopifnot(length(unique(cd4_preliminary_comparison$Patient_ID)) == 9)
```

```{r}
cd4_preliminary_comparison %>% filter(Level == "L2") %>%
  mutate(group = if_else(L2 == "Naive", "Naive", "Other")) %>%
  ggplot(aes(x = prop_all_preliminary, y = prop_all_final)) + 
  geom_smooth(method = "lm", formula = y ~ x) + 
  geom_point(aes(color = L2)) +
  facet_wrap(~group, scale = "free")

```

```{r}
cd4_preliminary_comparison_l2 <- cd4_preliminary_comparison %>% 
  filter(Level == "L2") %>%
  group_by(Patient_ID, Condition) %>%
  mutate(fold_change = ((n_final + 1) / sum_all_final) /   ((n_preliminary + 1) / sum_all_preliminary),
         perturbation_change = fold_change / sum(fold_change)
         #alr_perturbation_change = log(perturbation_change / perturbation_change[L2 == "Naive"]),
         #clr_perturbation_change = log(perturbation_change / exp(mean(log(perturbation_change))))
         )

cd4_preliminary_comparison_l2 %>%
  ggplot(aes(x = L2, y = fold_change, color = Condition)) + 
  geom_hline(yintercept = 1, color = "orangered") +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.8) +
  scale_y_log10() + coord_flip()
```

```{r}
cd4_preliminary_comparison_l2 %>%
  ggplot(aes(x = L2, y = perturbation_change, color = Condition)) + 
  geom_hline(yintercept = 1/length(unique(cd4_preliminary_comparison_l2$L2)), color = "orangered") +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.8) +
  scale_y_log10() + coord_flip()
```

```{r}
cd4_preliminary_comparison_l3 <- cd4_preliminary_comparison %>% 
  filter(Level == "L3") %>%
  group_by(Patient_ID, Condition) %>%
  mutate(fold_change = ((n_final + 1) / sum_all_final) /   ((n_preliminary + 1) / sum_all_preliminary),
         perturbation_change = fold_change / sum(fold_change))

cd4_preliminary_comparison_l3 %>%
  ggplot(aes(x = L3, y = fold_change, color = Condition)) + 
  geom_hline(yintercept = 1, color = "orangered") +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.8) +
  scale_y_log10()  + coord_flip()


```

```{r}
cd4_preliminary_comparison_l3 %>%
  ggplot(aes(x = L3, y = perturbation_change, color = Condition)) + 
  geom_hline(yintercept = 1/length(unique(cd4_preliminary_comparison_l3$L3)), color = "orangered") +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.8) +
  scale_y_log10() + coord_flip()
```

```{r}
# cd4_all_prop_with_preliminary %>% group_by(Patient_ID, Condition, isPreliminary) %>% summarise(n_samps = length(unique(Sample_ID))) %>%
#   filter(n_samps > 1)
# 
# cd4_all_prop_with_preliminary %>% group_by(Sample_ID) %>% summarise(n_samps = length(unique(paste0(Patient_ID, Condition, isPreliminary, sep = "___")))) %>%
#   filter(n_samps > 1)


cd4_for_calibration_model <- cd4_all_prop_with_preliminary %>%
  mutate(PatientDesc = paste(Patient_ID, Condition, sep = "___"),
         isFinal = as.numeric(!isPreliminary)) %>%
  group_by(PatientDesc) %>%
  filter(any(isPreliminary)) %>%
  ungroup() 

# fit_cd4_calibration <- brm(bf(n ~ offset(-log(sum_all / 1000)) + PatientDesc*L3 + isFinal * L3, 
#                   shape ~ L2 + (1 | L3)), 
#                data = cd4_for_calibration_model, 
#                control = list(adapt_delta = 0.95),
#                family = negbinomial(),
#                file = here::here("stored_fits/cd4_calibration.rds"),
#                file_refit = "on_change")
```


```{r}
cd4_test <- cd4_for_calibration_model %>% filter(L3 == "Naive1", Level == "L3") %>% 
  uncertainty_fold_change_group()
```


```{r}
#cd4_test <- cd4_for_calibration_model %>% filter(L2 == "Unconventional", Level == "L2")
cd4_test <- cd4_for_calibration_model %>% filter(L3 == "Naive1", Level == "L3")
cd4_test <- cd4_test %>% mutate(AgeStd = (10 - Age)/5)
fit_calibration <- MASS::glm.nb(n ~ offset(-log(sum_all / 1000)) + PatientDesc + isFinal, data = cd4_test)
plot(fit_calibration)
confint(fit_calibration)
fit_calibration$theta
fit_calibration$SE.theta

cd4_test %>% mutate(fitted = fitted(fit_calibration), residuals = residuals(fit_calibration)) %>%
  ggplot(aes(x = fitted, y = residuals, group = PatientDesc))  + geom_line() + geom_point(aes(color = as.factor(isFinal))) + scale_x_log10()

cd4_test %>% mutate(fitted = log(fitted(fit_calibration)), residuals = residuals(fit_calibration)) %>%
  ggplot(aes(x = fitted, y = log(n) - log(sum_all / 1000), group = PatientDesc))  + geom_abline(slope = 1, intercept = 0, color = "lightblue", linewidth = 2) + geom_line() + geom_point(aes(color = as.factor(isFinal))) 


cd4_test %>% mutate(fitted = log(fitted(fit_calibration)), residuals = residuals(fit_calibration)) %>%
  filter(Condition == "Ctrl T0") %>%
  ggplot(aes(x = fitted, y = log(n) - log(sum_all / 1000), group = PatientDesc))  + geom_abline(slope = 1, intercept = 0, color = "lightblue", linewidth = 2) + geom_line(aes(color = PatientDesc)) #+ geom_point(aes(color = as.factor(isFinal))) 

```

```{r}
cd4_test %>% mutate(fitted = log(fitted(fit_calibration)), residuals = residuals(fit_calibration)) %>%
  filter(Condition == "Ctrl T0")
```

```{r}
library(rstanarm)
stan_test <- rstanarm::stan_glm.nb(n ~ offset(-log(sum_all / 1000)) + PatientDesc + isFinal + isFinal:Condition + isFinal:AgeStd, data = cd4_test)

stan_test_df <- tidybayes::add_predicted_rvars(cd4_test, stan_test, offset = -log(cd4_test$sum_all / 1000)) %>%
  mutate(pred_ratio = (.prediction +  1) / sum_all) %>%
  select(n, Condition, sum_all, isPreliminary, PatientDesc, Sample_ID, Patient_ID, pred_ratio)  %>%
  mutate(isPreliminary = if_else(isPreliminary, "preliminary", "final")) %>%
  pivot_wider(names_from = isPreliminary, values_from = c("Sample_ID", "n", "sum_all", "pred_ratio")) %>%
  mutate(fold_change = ((n_final + 1) / sum_all_final) /   ((n_preliminary + 1) / sum_all_preliminary),
         fc_rvar = pred_ratio_final / pred_ratio_preliminary,
         low = as.numeric(quantile(fc_rvar, 0.025)), 
         high = as.numeric(quantile(fc_rvar, 0.975)))

stan_test_df %>%
  ggplot(aes(x = PatientDesc, y = fold_change, color = Condition)) + 
  geom_hline(yintercept = exp(fit_calibration$coefficients["isFinal"]), color = "blue") +
  geom_hline(yintercept = 1, color = "orangered") +
  geom_linerange(aes(ymin = low, ymax = high), alpha = 0.8) +
  geom_point(alpha = 0.8) +
  scale_y_log10() + coord_flip()
```

```{r}
summary(stan_test)
stan_test_df %>% filter(Patient_ID == 209) %>% pull(pred_ratio_preliminary) %>% posterior::draws_of() %>% hist()
stan_test_df %>% filter(Patient_ID == 209) %>% pull(pred_ratio_final) %>% posterior::draws_of() %>% hist()

stan_test %>% posterior::as_draws_rvars() %>% posterior::variables()
hist(1 / (stan_test %>% posterior::as_draws_rvars())$reciprocal_dispersion %>% posterior::draws_of())
```


```{r}
cc <- coefficients(fit_calibration)
base <- cc["(Intercept)"] + cc["PatientDesc209___Ctrl T0"]
base_f <- base + cc["isFinal"]
exp(base) / 1000
exp(base_f) / 1000
uncertainty_fold_change_single(5.942,1316, 5.759, 611, fit_calibration$theta, fit_calibration$SE.theta)
```


```{r}
hist(fold_change_with_uncertainty$sum_all_final / fold_change_with_uncertainty$sum_all_preliminary)
```


```{r}
cd4_test
```


## CD8

```{r}
cd8_all <-
  read_csv(
    here::here("private_data/all_levels_counts_cd8.csv"),
    col_types = cols(
      Sample_ID = col_character(),
      annotations = col_character(),
      n = col_integer(),
      Condition = col_character(),
      Condition2 = col_character(),
      Disease = col_character(),
      Sex = col_character(),
      Age = col_double(),
      Age_group = col_character(),
      Patient_ID = col_integer(),
      Time = col_character(),
      Experiment_ID = col_character(),
      Level = col_character()
    )
  ) %>%
  separate(annotations, into = c("L1", "L2", "L3", "Marker_genes"), sep = "---|: ", fill = "right") %>%
  mutate(Name = case_when(!is.na(L3) ~ L3,
                          !is.na(L2) ~ L2,
                          TRUE ~ L1))
```


```{r}
cd8_all_prop <- cd8_all %>% 
  group_by(Sample_ID, Level) %>%
  mutate(sum_parent = if_else(Level == "L1", sum(n), NA_real_),
         sum_all = sum(n)) %>%
  group_by(Sample_ID, L1, Level) %>%
  mutate(sum_parent = if_else(Level == "L2", sum(n), sum_parent)) %>%
  group_by(Sample_ID, L1, L2, Level) %>%
  mutate(sum_parent = if_else(Level == "L3", sum(n), sum_parent)) %>%
  ungroup() %>%
  mutate(prop_parent = n / sum_parent,
         prop_all = n / sum_all,
         parent_level = case_when(Level == "L1" ~ "All",
                                  Level == "L2" ~ L1,
                                  Level == "L3" ~ L2))

stopifnot(all(!is.na(cd8_all_prop$sum_parent)))
stopifnot(all(!is.na(cd8_all_prop$parent_level)))

prop_sums_bad <- cd8_all_prop %>%
  group_by(Sample_ID, Level, parent_level) %>%
  summarise(prop_sum = sum(prop_parent), .groups = "drop") %>%
  filter(abs(prop_sum - 1) > 1e-5)

stopifnot(nrow(prop_sums_bad) == 0)

sums_all_bad <- cd8_all_prop %>%
  group_by(Sample_ID, sum_all) %>%
  tally() %>%
  group_by(Sample_ID) %>%
  filter(n() > 1)

stopifnot(nrow(sums_all_bad) == 0)


```

```{r}
cd8_all_prop %>% filter(Level == "L1") %>% plot_prop()
cd8_all_prop %>% filter(Level == "L2" & L2 != "NK cells") %>% plot_prop()
```


```{r, fig.width=8, fig.height=8}
cd8_all_prop %>% filter(Level == "L3") %>% plot_prop()
```

Samples with extreme proportion of MAIT4:

```{r}
cd8_all_prop %>% filter(Name == "MAIT4", prop_parent > 0.3) %>% df_for_reading()
```


```{r}
cd8_all_prop %>% filter(Level == "L1") %>% plot_prop(prop_all)
cd8_all_prop %>% filter(Level == "L2" & L2 != "Unconventional") %>% plot_prop(prop_all)
```




```{r, fig.width=8, fig.height=8}
cd8_all_prop %>% filter(Level == "L3") %>% plot_prop(prop_all)
```

```{r}
cd8_l3 <- cd8_all_prop %>%
  filter(Level == "L3") %>%
  select(-Level) %>%
  mutate()

```


```{r cells-per-sample-cd8, fig.cap="Distribution of cells per sample in the cd8 data"}
cd8_l3 %>% group_by(Sample_ID) %>%
  summarise(n_cells = sum(n)) %>%
  ggplot(aes(n_cells)) + geom_histogram(binwidth = 100)
```

## Modelling

### CD4 - outliers removed

Those samples are claimed as outliers (proportion of the cells across the whole sample is 3 SD or more above the mean proportion across the given L3 population)

```{r}
remove_criteria <- function(x) {
  return(x - mean(x) > 3 * sd(x))
}


cd4_l3_for_model_alldata <- cd4_l3 %>% 
  mutate(isT1 = as.numeric(Condition == "Dia T1"), isDia = as.numeric(Disease == "Dia"))

cd4_l3_for_model_alldata %>% group_by(L3) %>%
  filter(remove_criteria(prop_all)) %>%
  df_for_reading() %>%
  arrange(L3)
```


```{r}


cd4_l3_for_model <- cd4_l3_for_model_alldata %>% 
  filter(!remove_criteria(prop_all))

fit_cd4 <- brm(bf(n ~ offset(-log(sum_all / 1000)) + isDia + isT1 + L3 + (0 + isDia + isT1 | L2) + (0 + isDia + isT1 | L3), 
                  shape ~ L2 + (1 | L3)), 
               data = cd4_l3_for_model, 
               control = list(adapt_delta = 0.95),
               iter = hmc_iters,
               family = negbinomial(),
               file = here::here("stored_fits/cd4.rds"),
               file_refit = "on_change")
```


```{r}
# make_stancode(bf(n ~ offset(-log(n_all / 1000)) + isDia + isT1 + L3 + (0 + isDia + isT1 | L2) + (0 + isDia + isT1 | L3), 
#                   shape ~ (1 | L3)), 
#                data = cd4_l3_for_model, 
#                family = negbinomial())
# 
# make_standata(bf(n ~ offset(-log(n_all / 1000)) + isDia + isT1 + L3 + (0 + isDia + isT1 | L2) + (0 + isDia + isT1 | L3), 
#                   shape ~ (1 | L3)), 
#                data = cd4_l3_for_model, 
#                family = negbinomial())$offsets

```


```{r}
fit_cd4
```

```{r}
#pp_check(fit_cd4, type = "loo_pit_overlay", ndraws = 50)
```

```{r}
if(do_ppc_checks) {
  pred_cd4 <- posterior_predict(fit_cd4)
  ppc_stat_grouped(cd4_l3_for_model$n, pred_cd4, group = cd4_l3_for_model$L2, stat = "mean")
  ppc_stat_grouped(cd4_l3_for_model$n, pred_cd4, group = cd4_l3_for_model$L2, stat = "sd")
  ppc_stat_grouped(cd4_l3_for_model$n, pred_cd4, group = cd4_l3_for_model$L2, stat = "phano")
  ppc_stat_grouped(cd4_l3_for_model$n, pred_cd4, group = cd4_l3_for_model$L2, stat = "prop_zero")
}

```

```{r}
if(do_ppc_checks) {
  ppc_stat_grouped(cd4_l3_for_model$n, pred_cd4, group = cd4_l3_for_model$L3, stat = "mean")
  ppc_stat_grouped(cd4_l3_for_model$n, pred_cd4, group = cd4_l3_for_model$L3, stat = "sd")
  ppc_stat_grouped(cd4_l3_for_model$n, pred_cd4, group = cd4_l3_for_model$L3, stat = "phano")
}
```



```{r}
if(do_loo) {
  loo_cd4 <- loo(fit_cd4)
}
```

List of highly influential observations (for the fit after removing outliers mentioned above, i.e. those are potentially still outliers):

```{r}
if(do_loo) {
  cd4_l3_for_model[loo_cd4$diagnostics$pareto_k > 0.5,] %>% df_for_reading()
}
```



### CD4 - full data

```{r}

fit_cd4_alldata <- brm(bf(n ~ offset(-log(sum_all / 1000)) + isDia + isT1 + L3 + (0 + isDia + isT1 | L2) + (0 + isDia + isT1 | L3), 
                  shape ~ L2 + (1 | L3)), 
               data = cd4_l3_for_model_alldata, 
               control = list(adapt_delta = 0.95),
               iter = hmc_iters,
               family = negbinomial(),
               file = here::here("stored_fits/cd4_alldata.rds"),
               file_refit = "on_change")
```


```{r}
fit_cd4_alldata
```

```{r}
if(do_ppc_checks) {

  pred_alldata <- posterior_predict(fit_cd4_alldata)
  
  ppc_stat_grouped(cd4_l3_for_model_alldata$n, pred_alldata, group = cd4_l3_for_model_alldata$L2, stat = "mean")
  ppc_stat_grouped(cd4_l3_for_model_alldata$n, pred_alldata, group = cd4_l3_for_model_alldata$L2, stat = "sd")
  ppc_stat_grouped(cd4_l3_for_model_alldata$n, pred_alldata, group = cd4_l3_for_model_alldata$L2, stat = "phano")
  ppc_stat_grouped(cd4_l3_for_model_alldata$n, pred_alldata, group = cd4_l3_for_model_alldata$L2, stat = "prop_zero")
}
#ppc_bars_grouped(cd4_l3_for_model$n, pred, group = cd4_l3_for_model$L2)
```
```{r}
if(do_ppc_checks) {
  ppc_stat_grouped(cd4_l3_for_model_alldata$n, pred, group = cd4_l3_for_model_alldata$L3, stat = "sd")
  ppc_stat_grouped(cd4_l3_for_model_alldata$n, pred, group = cd4_l3_for_model_alldata$L3, stat = "phano")
}
```


### CD4 - idaa1c

Using idaac to predict populations at time 0 to see potential differences predicting remission

```{r}
cd4_l3_for_model_idaa1c <- cd4_l3_for_model_alldata %>%
  filter(Time == "T0") %>%
  inner_join(clinical_data %>% filter(Time == "T0") , by = "Patient_ID", 
             relationship = "many-to-one",
             unmatched = "error") %>%
  filter(Condition == "Dia T0", !is.na(idaa1c)) %>%
  mutate(idaa1c = idaa1c - mean(idaa1c))

fit_cd4_idaa1c <- brm(bf(n ~ offset(-log(sum_all / 1000)) + idaa1c + L3 + (0 + idaa1c | L2) + (0 + idaa1c | L3), 
                  shape ~ L2 + (1 | L3)), 
               data = cd4_l3_for_model_idaa1c, 
               control = list(adapt_delta = 0.95),
               iter = hmc_iters,
               family = negbinomial(),
               file = here::here("stored_fits/cd4_idaa1c.rds"),
               file_refit = "on_change")

fit_cd4_idaa1c
```


### CD8 - outliers removed

The following are claimed as outliers:

```{r}
cd8_l3_for_model_alldata <- cd8_l3 %>% 
  mutate(isT1 = as.numeric(Condition == "Dia T1"), isDia = as.numeric(Disease == "Dia"))


cd8_l3_for_model_alldata %>% group_by(L3) %>%
  filter(remove_criteria(prop_all)) %>%
  df_for_reading() %>%
  arrange(L3)

```


```{r}

cd8_l3_for_model <- cd8_l3_for_model_alldata %>% 
  filter(!remove_criteria(prop_all))

fit_cd8 <- brm(bf(n ~ offset(-log(sum_all / 1000)) + isDia + isT1 + L3 + (0 + isDia + isT1 | L2) + (0 + isDia + isT1 | L3), 
                  shape ~ L2 + (1 | L3)), 
               data = cd8_l3_for_model, 
               family = negbinomial(),
               control = list(adapt_delta = 0.95),
               iter = hmc_iters,
               file = here::here("stored_fits/cd8.rds"),
               file_refit = "on_change")
```


```{r}
fit_cd8
```

```{r}
#pp_check(fit_cd8, type = "loo_pit_overlay", ndraws = 50)
```

```{r}
if(do_ppc_checks) {
  pred_cd8 <- posterior_predict(fit_cd8)
  
  ppc_stat_grouped(cd8_l3_for_model$n, pred_cd8, group = cd8_l3_for_model$L2, stat = "mean")
  ppc_stat_grouped(cd8_l3_for_model$n, pred_cd8, group = cd8_l3_for_model$L2, stat = "sd")
  ppc_stat_grouped(cd8_l3_for_model$n, pred_cd8, group = cd8_l3_for_model$L2, stat = "phano")
  ppc_stat_grouped(cd8_l3_for_model$n, pred_cd8, group = cd8_l3_for_model$L2, stat = "prop_zero")
}
```
```{r}
if(do_ppc_checks) {
  ppc_stat_grouped(cd8_l3_for_model$n, pred_cd8, group = cd8_l3_for_model$L3, stat = "mean")
  ppc_stat_grouped(cd8_l3_for_model$n, pred_cd8, group = cd8_l3_for_model$L3, stat = "sd")
  ppc_stat_grouped(cd8_l3_for_model$n, pred_cd8, group = cd8_l3_for_model$L3, stat = "phano")
}
```



```{r}
if(do_loo) {
  loo_cd8 <- loo(fit_cd8)
}
```

List of highly influential observation (with outliers removed, i.e. potential extra outliers)

```{r}
if(do_loo) {
  cd8_l3_for_model[loo_cd8$diagnostics$pareto_k > 0.5,] %>% df_for_reading()
}
```



### CD8 - full data

```{r}

fit_cd8_alldata <- brm(bf(n ~ offset(-log(sum_all / 1000)) + isDia + isT1 + L3 + (0 + isDia + isT1 | L2) + (0 + isDia + isT1 | L3), 
                  shape ~ L2 + (1 | L3)), 
               data = cd8_l3_for_model_alldata, 
               control = list(adapt_delta = 0.95),
               iter = hmc_iters,
               family = negbinomial(),
               file = here::here("stored_fits/cd8_alldata.rds"),
               file_refit = "on_change")
```


```{r}
fit_cd8_alldata
```



```{r}
#pp_check(fit_cd8_alldata, type = "loo_pit_overlay", ndraws = 50)
```

```{r}
if(do_ppc_checks) {

  pred_alldata <- posterior_predict(fit_cd8_alldata)
  
  ppc_stat_grouped(cd8_l3_for_model_alldata$n, pred_alldata, group = cd8_l3_for_model_alldata$L2, stat = "mean")
  ppc_stat_grouped(cd8_l3_for_model_alldata$n, pred_alldata, group = cd8_l3_for_model_alldata$L2, stat = "sd")
  ppc_stat_grouped(cd8_l3_for_model_alldata$n, pred_alldata, group = cd8_l3_for_model_alldata$L2, stat = "phano")
  ppc_stat_grouped(cd8_l3_for_model_alldata$n, pred_alldata, group = cd8_l3_for_model_alldata$L2, stat = "prop_zero")
  
  #ppc_bars_grouped(cd8_l3_for_model$n, pred, group = cd8_l3_for_model$L2)
}
```
```{r}
if(do_ppc_checks) {
  ppc_stat_grouped(cd8_l3_for_model_alldata$n, pred_alldata, group = cd8_l3_for_model_alldata$L3, stat = "sd")
  ppc_stat_grouped(cd8_l3_for_model_alldata$n, pred_alldata, group = cd8_l3_for_model_alldata$L3, stat = "phano")
}
```


### CD8 - idaa1c

Using idaac to predict populations at time 0 to see potential differences predicting remission

```{r}
cd8_l3_for_model_idaa1c <- cd8_l3_for_model_alldata %>%
  filter(Time == "T0") %>%
  inner_join(clinical_data %>% filter(Time == "T0") , by = "Patient_ID", 
             relationship = "many-to-one",
             unmatched = "error") %>%
  filter(Condition == "Dia T0", !is.na(idaa1c))  %>%
  mutate(idaa1c = idaa1c - mean(idaa1c))

fit_cd8_idaa1c <- brm(bf(n ~ offset(-log(sum_all / 1000)) + idaa1c + L3 + (0 + idaa1c | L2) + (0 + idaa1c | L3), 
                  shape ~ L2 + (1 | L3)), 
               data = cd8_l3_for_model_idaa1c, 
               control = list(adapt_delta = 0.95),
               iter = hmc_iters,
               family = negbinomial(),
               file = here::here("stored_fits/cd8_idaa1c.rds"),
               file_refit = "on_change")

fit_cd8_idaa1c
```


### Associations summary

In all cases, the isDia estimates show the difference (fold change in abundance) between healthy controls and diabetics at time 0 (fold change > 1 means higher abundance in diabetics). The isT1 estimates are the difference for diabetics between time 1 and and time 0 (fold change > 1 means higher abundance at time 1).

We look separately at L2 populations relative to total, L3 populations relative to parent population and then at the combined estimate - L3 populations relative to total.

Finally we look at the dataset with outliers removed (proportion relative to total is 3SD or more above others) and the full dataset without any removals.

```{r, fig.height=2.8, fig.width=8}
plot_population_effects(get_l2_draws(fit_cd4) %>% add_diaT1_draws(), "CD4 - L2 relative to total")
plot_population_effects(get_l2_draws(fit_cd4_alldata) %>% add_diaT1_draws(), "CD4 - L2 relative to total, no removals")
plot_population_effects(get_l2_draws(fit_cd4_idaa1c), "CD4 - L2 relative to total, remission")
```

```{r, fig.width=8, fig.height=7}
plot_population_effects(get_l3_draws(fit_cd4) %>% add_diaT1_draws(), "CD4 - L3 relative to parent")
plot_population_effects(get_l3_draws(fit_cd4_alldata) %>% add_diaT1_draws(), "CD4 - L3 relative to parent, no removals")
plot_population_effects(get_l3_draws(fit_cd4_idaa1c), "CD4 - L3 relative to parent, remission")

plot_population_effects(get_combined_draws(fit_cd4, cd4_l3_for_model) %>% add_diaT1_draws(), "CD4 - L3 relative to total (combined)")
plot_population_effects(get_combined_draws(fit_cd4_alldata, cd4_l3_for_model_alldata) %>% add_diaT1_draws(), "CD4 - L3 relative to total (combined), no removals")
plot_population_effects(get_combined_draws(fit_cd4_idaa1c, cd4_l3_for_model_idaa1c), "CD4 - L3 relative to parent, remission")

```



```{r, fig.height=2.5, fig.width=8}
plot_population_effects(get_l2_draws(fit_cd8) %>% add_diaT1_draws(), "CD8 - L2 relative to total")
plot_population_effects(get_l2_draws(fit_cd8_alldata) %>% add_diaT1_draws(), "CD8 - L2 relative to total, no removals")
plot_population_effects(get_l2_draws(fit_cd8_idaa1c), "CD8 - L2 relative to total, remission")
```
```{r, fig.width=8, fig.height=7.5}
plot_population_effects(get_l3_draws(fit_cd8) %>% add_diaT1_draws(), "CD8 - L3 relative to parent")
plot_population_effects(get_l3_draws(fit_cd8_alldata) %>% add_diaT1_draws(), "CD8 - L3 relative to parent, no removals")
plot_population_effects(get_l3_draws(fit_cd8_idaa1c), "CD8 - L3 relative to parent, remission")

plot_population_effects(get_combined_draws(fit_cd8, cd8_l3_for_model) %>% add_diaT1_draws(), "CD8 -  L3 relative to total (combined)")
plot_population_effects(get_combined_draws(fit_cd8_alldata, cd8_l3_for_model_alldata) %>% add_diaT1_draws(), "CD8 -  L3 relative to total (combined), no removals")
plot_population_effects(get_combined_draws(fit_cd8_idaa1c, cd8_l3_for_model_idaa1c), "CD8 - L3 relative to total, remission")
```



