---
title: "Bayes - Diabetes"
output: 
  word_document:
    toc: false
    reference_docx: "output_template.docx"
    df_print: "kable"
date: "Version `r format(Sys.time(), '%Y-%m-%d')`"

---


```{r setup, echo=FALSE, include=FALSE}
library(tidyverse)
library(brms)
library(bayesplot)
options(mc.cores = 4, brms.backend = "cmdstanr")
devtools::load_all()
theme_set(theme_minimal())
knitr::opts_chunk$set(echo = FALSE)

hmc_iters <- 7000
do_ppc_checks <- FALSE
```


```{r}
cd4_all <-
  read_csv(
    here::here("private_data/all_levels_counts_cd4.csv"),
    col_types = cols(
      Sample_ID = col_character(),
      annotations = col_character(),
      n = col_integer(),
      Condition = col_character(),
      Condition2 = col_character(),
      Disease = col_character(),
      Sex = col_character(),
      Age = col_double(),
      Age_group = col_character(),
      Patient_ID = col_integer(),
      Time = col_character(),
      Experiment_ID = col_character(),
      Level = col_character()
    )
  ) %>%
  separate(annotations, into = c("L1", "L2", "L3", "Marker_genes"), sep = "---|: ", fill = "right") %>%
  mutate(Name = case_when(!is.na(L3) ~ L3,
                          !is.na(L2) ~ L2,
                          TRUE ~ L1))
```


```{r}
cd4_all_prop <- cd4_all %>% 
  group_by(Sample_ID, Level) %>%
  mutate(sum_parent = if_else(Level == "L1", sum(n), NA_real_),
         sum_all = sum(n)) %>%
  group_by(Sample_ID, L1, Level) %>%
  mutate(sum_parent = if_else(Level == "L2", sum(n), sum_parent)) %>%
  group_by(Sample_ID, L1, L2, Level) %>%
  mutate(sum_parent = if_else(Level == "L3", sum(n), sum_parent)) %>%
  ungroup() %>%
  mutate(prop_parent = n / sum_parent,
         prop_all = n / sum_all,
         parent_level = case_when(Level == "L1" ~ "All",
                                  Level == "L2" ~ L1,
                                  Level == "L3" ~ L2))

stopifnot(all(!is.na(cd4_all_prop$sum_parent)))
stopifnot(all(!is.na(cd4_all_prop$parent_level)))

prop_sums_bad <- cd4_all_prop %>%
  group_by(Sample_ID, Level, parent_level) %>%
  summarise(prop_sum = sum(prop_parent), .groups = "drop") %>%
  filter(abs(prop_sum - 1) > 1e-5)

stopifnot(nrow(prop_sums_bad) == 0)

sums_all_bad <- cd4_all_prop %>%
  group_by(Sample_ID, sum_all) %>%
  tally() %>%
  group_by(Sample_ID) %>%
  filter(n() > 1)

stopifnot(nrow(sums_all_bad) == 0)


```

## CD4

```{r}
plot_prop <- function(all_prop, prop_col = prop_parent) {
  all_prop %>%
    filter(!is.na({{prop_col}})) %>%
    ggplot(aes(x = Condition, y = {{prop_col}})) +
    geom_point(position = position_jitter(width = 0.3, height = 0), alpha = 0.5, color = "darkgray") +
    stat_summary(fun.data = "mean_se") +
    facet_wrap(~Name, scales = "free_y")
  
}
cd4_all_prop %>% filter(Level == "L1") %>% plot_prop()
cd4_all_prop %>% filter(Level == "L2" & L2 != "Unconventional") %>% plot_prop()
```


```{r, fig.width=8, fig.height=8}
cd4_all_prop %>% filter(Level == "L3") %>% plot_prop()
```

```{r}
cd4_all_prop %>% filter(Level == "L1") %>% plot_prop(prop_all)
cd4_all_prop %>% filter(Level == "L2" & L2 != "Unconventional") %>% plot_prop(prop_all)
```

```{r}
#cd4_all_prop %>% filter(Level == "L2" & L2 != "Unconventional") %>% 
  
```



```{r, fig.width=8, fig.height=8}
cd4_all_prop %>% filter(Level == "L3") %>% plot_prop(prop_all)
```

Samples with extreme proportions of ISAGhi:

```{r}
df_for_reading <- function(df) {
  df %>% select(Patient_ID, Sample_ID, L2, L3, Condition, Experiment_ID, n, prop_parent, prop_all) %>% mutate(across(starts_with("prop"), scales::percent)) %>%
    rename(Pat. = Patient_ID, Samp. = Sample_ID, Cond. = Condition, Exp = Experiment_ID, `% parent` = prop_parent, `% all` = prop_all)
}

cd4_all_prop %>% filter(Name == "ISAGhi" & prop_parent > 0.1) %>% df_for_reading() %>% select(-L3)
```


```{r}
cd4_l3 <- cd4_all_prop %>%
  filter(Level == "L3") %>%
  select(-Level) 
```


```{r cells-per-sample, fig.cap="Distribution of cells per sample in the CD4 data"}
cd4_l3 %>% group_by(Sample_ID) %>%
  summarise(n_cells = sum(n)) %>%
  ggplot(aes(n_cells)) + geom_histogram(binwidth = 100)
```

## CD8

```{r}
cd8_all <-
  read_csv(
    here::here("private_data/all_levels_counts_cd8.csv"),
    col_types = cols(
      Sample_ID = col_character(),
      annotations = col_character(),
      n = col_integer(),
      Condition = col_character(),
      Condition2 = col_character(),
      Disease = col_character(),
      Sex = col_character(),
      Age = col_double(),
      Age_group = col_character(),
      Patient_ID = col_integer(),
      Time = col_character(),
      Experiment_ID = col_character(),
      Level = col_character()
    )
  ) %>%
  separate(annotations, into = c("L1", "L2", "L3", "Marker_genes"), sep = "---|: ", fill = "right") %>%
  mutate(Name = case_when(!is.na(L3) ~ L3,
                          !is.na(L2) ~ L2,
                          TRUE ~ L1))
```


```{r}
cd8_all_prop <- cd8_all %>% 
  group_by(Sample_ID, Level) %>%
  mutate(sum_parent = if_else(Level == "L1", sum(n), NA_real_),
         sum_all = sum(n)) %>%
  group_by(Sample_ID, L1, Level) %>%
  mutate(sum_parent = if_else(Level == "L2", sum(n), sum_parent)) %>%
  group_by(Sample_ID, L1, L2, Level) %>%
  mutate(sum_parent = if_else(Level == "L3", sum(n), sum_parent)) %>%
  ungroup() %>%
  mutate(prop_parent = n / sum_parent,
         prop_all = n / sum_all,
         parent_level = case_when(Level == "L1" ~ "All",
                                  Level == "L2" ~ L1,
                                  Level == "L3" ~ L2))

stopifnot(all(!is.na(cd8_all_prop$sum_parent)))
stopifnot(all(!is.na(cd8_all_prop$parent_level)))

prop_sums_bad <- cd8_all_prop %>%
  group_by(Sample_ID, Level, parent_level) %>%
  summarise(prop_sum = sum(prop_parent), .groups = "drop") %>%
  filter(abs(prop_sum - 1) > 1e-5)

stopifnot(nrow(prop_sums_bad) == 0)

sums_all_bad <- cd8_all_prop %>%
  group_by(Sample_ID, sum_all) %>%
  tally() %>%
  group_by(Sample_ID) %>%
  filter(n() > 1)

stopifnot(nrow(sums_all_bad) == 0)


```

```{r}
cd8_all_prop %>% filter(Level == "L1") %>% plot_prop()
cd8_all_prop %>% filter(Level == "L2" & L2 != "NK cells") %>% plot_prop()
```


```{r, fig.width=8, fig.height=8}
cd8_all_prop %>% filter(Level == "L3") %>% plot_prop()
```

Samples with extreme proportion of MAIT4:

```{r}
cd8_all_prop %>% filter(Name == "MAIT4", prop_parent > 0.3) %>% df_for_reading()
```


```{r}
cd8_all_prop %>% filter(Level == "L1") %>% plot_prop(prop_all)
cd8_all_prop %>% filter(Level == "L2" & L2 != "Unconventional") %>% plot_prop(prop_all)
```




```{r, fig.width=8, fig.height=8}
cd8_all_prop %>% filter(Level == "L3") %>% plot_prop(prop_all)
```

```{r}
cd8_l3 <- cd8_all_prop %>%
  filter(Level == "L3") %>%
  select(-Level) %>%
  mutate()

```


```{r cells-per-sample-cd8, fig.cap="Distribution of cells per sample in the cd8 data"}
cd8_l3 %>% group_by(Sample_ID) %>%
  summarise(n_cells = sum(n)) %>%
  ggplot(aes(n_cells)) + geom_histogram(binwidth = 100)
```

## Modelling

### CD4 - outliers removed

Those samples are claimed as outliers (proportion of the cells across the whole sample is 3 SD or more above the mean proportion across the given L3 population)

```{r}
remove_criteria <- function(x) {
  return(x - mean(x) > 3 * sd(x))
}


cd4_l3_for_model_alldata <- cd4_l3 %>% 
  mutate(isT1 = as.numeric(Condition == "Dia T1"), isDia = as.numeric(Disease == "Dia"))

cd4_l3_for_model_alldata %>% group_by(L3) %>%
  filter(remove_criteria(prop_all)) %>%
  df_for_reading() %>%
  arrange(L3)
```


```{r}


cd4_l3_for_model <- cd4_l3_for_model_alldata %>% 
  filter(!remove_criteria(prop_all))

fit_cd4 <- brm(bf(n ~ offset(-log(sum_all / 1000)) + isDia + isT1 + L3 + (0 + isDia + isT1 | L2) + (0 + isDia + isT1 | L3), 
                  shape ~ L2 + (1 | L3)), 
               data = cd4_l3_for_model, 
               control = list(adapt_delta = 0.95),
               iter = hmc_iters,
               family = negbinomial(),
               file = here::here("stored_fits/cd4.rds"),
               file_refit = "on_change")
```


```{r}
# make_stancode(bf(n ~ offset(-log(n_all / 1000)) + isDia + isT1 + L3 + (0 + isDia + isT1 | L2) + (0 + isDia + isT1 | L3), 
#                   shape ~ (1 | L3)), 
#                data = cd4_l3_for_model, 
#                family = negbinomial())
# 
# make_standata(bf(n ~ offset(-log(n_all / 1000)) + isDia + isT1 + L3 + (0 + isDia + isT1 | L2) + (0 + isDia + isT1 | L3), 
#                   shape ~ (1 | L3)), 
#                data = cd4_l3_for_model, 
#                family = negbinomial())$offsets

```


```{r}
fit_cd4
```

```{r}
#pp_check(fit_cd4, type = "loo_pit_overlay", ndraws = 50)
```

```{r}
if(do_ppc_checks) {
  pred_cd4 <- posterior_predict(fit_cd4)
  ppc_stat_grouped(cd4_l3_for_model$n, pred_cd4, group = cd4_l3_for_model$L2, stat = "mean")
  ppc_stat_grouped(cd4_l3_for_model$n, pred_cd4, group = cd4_l3_for_model$L2, stat = "sd")
  ppc_stat_grouped(cd4_l3_for_model$n, pred_cd4, group = cd4_l3_for_model$L2, stat = "phano")
  ppc_stat_grouped(cd4_l3_for_model$n, pred_cd4, group = cd4_l3_for_model$L2, stat = "prop_zero")
}

```

```{r}
if(do_ppc_checks) {
  ppc_stat_grouped(cd4_l3_for_model$n, pred_cd4, group = cd4_l3_for_model$L3, stat = "mean")
  ppc_stat_grouped(cd4_l3_for_model$n, pred_cd4, group = cd4_l3_for_model$L3, stat = "sd")
  ppc_stat_grouped(cd4_l3_for_model$n, pred_cd4, group = cd4_l3_for_model$L3, stat = "phano")
}
```



```{r}
loo_cd4 <- loo(fit_cd4)
```

List of highly influential observations (for the fit after removing outliers mentioned above, i.e. those are potentially still outliers):

```{r}
cd4_l3_for_model[loo_cd4$diagnostics$pareto_k > 0.5,] %>% df_for_reading()
```



### CD4 - full data

```{r}

fit_cd4_alldata <- brm(bf(n ~ offset(-log(sum_all / 1000)) + isDia + isT1 + L3 + (0 + isDia + isT1 | L2) + (0 + isDia + isT1 | L3), 
                  shape ~ L2 + (1 | L3)), 
               data = cd4_l3_for_model_alldata, 
               control = list(adapt_delta = 0.95),
               iter = hmc_iters,
               family = negbinomial(),
               file = here::here("stored_fits/cd4_alldata.rds"),
               file_refit = "on_change")
```


```{r}
fit_cd4_alldata
```

```{r}
if(do_ppc_checks) {

  pred_alldata <- posterior_predict(fit_cd4_alldata)
  
  ppc_stat_grouped(cd4_l3_for_model_alldata$n, pred_alldata, group = cd4_l3_for_model_alldata$L2, stat = "mean")
  ppc_stat_grouped(cd4_l3_for_model_alldata$n, pred_alldata, group = cd4_l3_for_model_alldata$L2, stat = "sd")
  ppc_stat_grouped(cd4_l3_for_model_alldata$n, pred_alldata, group = cd4_l3_for_model_alldata$L2, stat = "phano")
  ppc_stat_grouped(cd4_l3_for_model_alldata$n, pred_alldata, group = cd4_l3_for_model_alldata$L2, stat = "prop_zero")
}
#ppc_bars_grouped(cd4_l3_for_model$n, pred, group = cd4_l3_for_model$L2)
```
```{r}
if(do_ppc_checks) {
  ppc_stat_grouped(cd4_l3_for_model_alldata$n, pred, group = cd4_l3_for_model_alldata$L3, stat = "sd")
  ppc_stat_grouped(cd4_l3_for_model_alldata$n, pred, group = cd4_l3_for_model_alldata$L3, stat = "phano")
}
```



### CD8 - outliers removed

The following are claimed as outliers:

```{r}
cd8_l3_for_model_alldata <- cd8_l3 %>% 
  mutate(isT1 = as.numeric(Condition == "Dia T1"), isDia = as.numeric(Disease == "Dia"))


cd8_l3_for_model_alldata %>% group_by(L3) %>%
  filter(remove_criteria(prop_all)) %>%
  df_for_reading() %>%
  arrange(L3)

```


```{r}

cd8_l3_for_model <- cd8_l3_for_model_alldata %>% 
  filter(!remove_criteria(prop_all))

fit_cd8 <- brm(bf(n ~ offset(-log(sum_all / 1000)) + isDia + isT1 + L3 + (0 + isDia + isT1 | L2) + (0 + isDia + isT1 | L3), 
                  shape ~ L2 + (1 | L3)), 
               data = cd8_l3_for_model, 
               family = negbinomial(),
               control = list(adapt_delta = 0.95),
               iter = hmc_iters,
               file = here::here("stored_fits/cd8.rds"),
               file_refit = "on_change")
```


```{r}
fit_cd8
```

```{r}
#pp_check(fit_cd8, type = "loo_pit_overlay", ndraws = 50)
```

```{r}
if(do_ppc_checks) {
  pred_cd8 <- posterior_predict(fit_cd8)
  
  ppc_stat_grouped(cd8_l3_for_model$n, pred_cd8, group = cd8_l3_for_model$L2, stat = "mean")
  ppc_stat_grouped(cd8_l3_for_model$n, pred_cd8, group = cd8_l3_for_model$L2, stat = "sd")
  ppc_stat_grouped(cd8_l3_for_model$n, pred_cd8, group = cd8_l3_for_model$L2, stat = "phano")
  ppc_stat_grouped(cd8_l3_for_model$n, pred_cd8, group = cd8_l3_for_model$L2, stat = "prop_zero")
}
```
```{r}
if(do_ppc_checks) {
  ppc_stat_grouped(cd8_l3_for_model$n, pred_cd8, group = cd8_l3_for_model$L3, stat = "mean")
  ppc_stat_grouped(cd8_l3_for_model$n, pred_cd8, group = cd8_l3_for_model$L3, stat = "sd")
  ppc_stat_grouped(cd8_l3_for_model$n, pred_cd8, group = cd8_l3_for_model$L3, stat = "phano")
}
```



```{r}
loo_cd8 <- loo(fit_cd8)
```

List of highly influential observation (with outliers removed, i.e. potential extra outliers)

```{r}
cd8_l3_for_model[loo_cd8$diagnostics$pareto_k > 0.5,] %>% df_for_reading()
```



### CD8 - full data

```{r}

fit_cd8_alldata <- brm(bf(n ~ offset(-log(sum_all / 1000)) + isDia + isT1 + L3 + (0 + isDia + isT1 | L2) + (0 + isDia + isT1 | L3), 
                  shape ~ L2 + (1 | L3)), 
               data = cd8_l3_for_model_alldata, 
               control = list(adapt_delta = 0.95),
               iter = hmc_iters,
               family = negbinomial(),
               file = here::here("stored_fits/cd8_alldata.rds"),
               file_refit = "on_change")
```


```{r}
fit_cd8_alldata
```



```{r}
#pp_check(fit_cd8_alldata, type = "loo_pit_overlay", ndraws = 50)
```

```{r}
if(do_ppc_checks) {

  pred_alldata <- posterior_predict(fit_cd8_alldata)
  
  ppc_stat_grouped(cd8_l3_for_model_alldata$n, pred_alldata, group = cd8_l3_for_model_alldata$L2, stat = "mean")
  ppc_stat_grouped(cd8_l3_for_model_alldata$n, pred_alldata, group = cd8_l3_for_model_alldata$L2, stat = "sd")
  ppc_stat_grouped(cd8_l3_for_model_alldata$n, pred_alldata, group = cd8_l3_for_model_alldata$L2, stat = "phano")
  ppc_stat_grouped(cd8_l3_for_model_alldata$n, pred_alldata, group = cd8_l3_for_model_alldata$L2, stat = "prop_zero")
  
  #ppc_bars_grouped(cd8_l3_for_model$n, pred, group = cd8_l3_for_model$L2)
}
```
```{r}
if(do_ppc_checks) {
  ppc_stat_grouped(cd8_l3_for_model_alldata$n, pred_alldata, group = cd8_l3_for_model_alldata$L3, stat = "sd")
  ppc_stat_grouped(cd8_l3_for_model_alldata$n, pred_alldata, group = cd8_l3_for_model_alldata$L3, stat = "phano")
}
```


### Associations summary

In all cases, the isDia estimates show the difference (fold change in abundance) between healthy controls and diabetics at time 0 (fold change > 1 means higher abundance in diabetics). The isT1 estimates are the difference for diabetics between time 1 and and time 0 (fold change > 1 means higher abundance at time 1).

We look separately at L2 populations relative to total, L3 populations relative to parent population and then at the combined estimate - L3 populations relative to total.

Finally we look at the dataset with outliers removed (proportion relative to total is 3SD or more above others) and the full dataset without any removals.

```{r, fig.height=2.8, fig.width=6}
plot_population_effects(get_l2_draws(fit_cd4), "CD4 - L2 relative to total")
plot_population_effects(get_l2_draws(fit_cd4_alldata), "CD4 - L2 relative to total, no removals")
```

```{r, fig.width=8, fig.height=7}
plot_population_effects(get_l3_draws(fit_cd4), "CD4 - L3 relative to parent")
plot_population_effects(get_l3_draws(fit_cd4_alldata), "CD4 - L3 relative to parent, no removals")

plot_population_effects(get_combined_draws(fit_cd4, cd4_l3_for_model), "CD4 - L3 relative to total (combined)")
plot_population_effects(get_combined_draws(fit_cd4_alldata, cd4_l3_for_model_alldata), "CD4 - L3 relative to total (combined), no removals")
```



```{r, fig.height=2.5, fig.width=6}
plot_population_effects(get_l2_draws(fit_cd8), "CD8 - L2 relative to total")
plot_population_effects(get_l2_draws(fit_cd8_alldata), "CD8 - L2 relative to total, no removals")
```
```{r, fig.width=8, fig.height=7.5}
plot_population_effects(get_l3_draws(fit_cd8), "CD8 - L3 relative to parent")
plot_population_effects(get_l3_draws(fit_cd8_alldata), "CD8 - L3 relative to parent, no removals")

plot_population_effects(get_combined_draws(fit_cd8, cd8_l3_for_model), "CD8 -  L3 relative to total (combined)")
plot_population_effects(get_combined_draws(fit_cd8_alldata, cd8_l3_for_model_alldata), "CD8 -  L3 relative to total (combined), no removals")
```



